The Spatial and Temporal Domains of Modern Ecology

### 2017-12-22 paper/main.tex
Edited affiliations and made a few wording/typo changes suggested by Paul

\author
{Lyndon Estes$^{\ast1, 2, 3}$, Paul R. Elsen$^{2, 4}$, Tim Treuer$^{5}$, Labeeb Ahmed$^{6}$, \\
Kelly Caylor$^{3, 7}$, Jason Chang$^{6}$, Jonathan J. Choi$^{5}$, and Erle Ellis$^{6}$ \\
\\
\normalsize{$^{1}$Graduate School of Geography, Clark University, Worcester MA 01610, USA}\\
\normalsize{$^{2}$Woodrow Wilson School, Princeton University, Princeton, NJ 08544, USA}\\
\normalsize{$^{3}$Civil and Environmental Engineering, Princeton University, Princeton, NJ 08544, USA}\\
\normalsize{$^{4}$Department of Environmental Science, Policy, and Management,}\\
\normalsize{University of California Berkeley, Berkeley, CA 94720, USA}\\
\normalsize{$^{5}$Ecology and Evolutionary Biology, Princeton University, Princeton, NJ 08544, USA}\\
\normalsize{$^{6}$Geography and Environmental Systems, University of Maryland Baltimore County,}\\ 
\normalsize{Baltimore, MD 21250, USA}\\
\normalsize{$^{7}$Earth Research Institute, University of California Santa Barbara, Santa Barbara, CA 93106, USA}\\
\\
\normalsize{$^\ast$To whom correspondence should be addressed; E-mail:  LEstes@clarku.edu.}}
% Include the date command, but leave its argument blank.

@@

\linenumbers
\vspace{10pt}
The scales at which ecosystems are observed plays a critical role in shaping our understanding of their structure and function \cite{levin_problem_1992,chave_problem_2013,wiens_spatial_1989}.  Ecological patterns emerge from temporal and spatial domains that may be coarser or finer than the processes that shape them, which means that investigation across multiple scales is essential for understanding ecological phenomena \cite{levin_problem_1992,sandel_scale_2009}. This awareness has grown rapidly since the 1980s, accelerated by the need to understand how changes in the global climate, ocean, and land systems are affecting everything from individual populations \cite{tingley_push_2012} to entire biomes \cite{xiao_photosynthetic_2004}, while technological advances in areas such as remote sensing and genetics are making it ever-easier to quantify ecological features across a broad and increasing range of scales \cite{schneider_rise_2001, chave_problem_2013}.

Given the growing awareness of scale, expanding data gathering capabilities, and the fact that the most comprehensive (and arguably best-known) meta-analyses \cite{tilman_ecological_1989,kareiva_spatial_1988} of ecological research scales were published nearly 30 years ago (but see \cite{porter_crop_2005,sandel_scale_2009} for more recent reviews), it is both timely and important to assess the scales of contemporary ecological investigation. To address this need, we quantified the spatial and temporal domains (here domain means the distribution of observations within the spectrum of one or more scale dimensions\footnote{This definition differs slightly from Wiens' \cite{wiens_spatial_1989}, who defined ``domain of scale'' as ``a portion of the scale spectrum within which process-pattern relationships are consistent regardless of scale.''}) of empirical observations (defined here as ecological observations collected under un-controlled or non-manipulated conditions) that were reported within recently (2004-2014) published ecological studies. Empirical observations are critical for developing and testing the models that explain why ecological patterns vary in time and space \cite{levin_problem_1992, tilman_ecological_1989}, therefore the spatio-temporal domains of observations provide an important indicator of the field's progress towards achieving a holistic, predictive understanding of ecosystems \cite{chave_problem_2013,levin_problem_1992}.

@@

Contrasting observational dimensions reveals that larger extents are associated with larger spatial replicates (Fig. 2C), while longer durations tend to be associated with longer intervals (Fig. 2B). The latter association primarily reflects a cost-imposed tradeoff between sampling frequency and temporal duration that is characteristic of traditional field-based observation, the dominant observational mode, but is also reflected by the relative domain locations of the four main observational methods, suggesting that high observational frequency poses a durational cost across observational methods. A similar cost tradeoff is also evident in for the inverse relationship between resolution and interval that dominates that domain space (Fig. 2A), which again primarily relates to characteristics of field-based observations, in which larger spatial replicates demand greater effort that in turn reduces sampling frequency \cite{kareiva_spatial_1988}. Less obvious is the opposite tradeoff that occurs with remote observation (but shown in Fig. S2 in SI), where finer resolution is desired to increase detail, but typically comes at the cost of longer intervals \cite{estes_platform_2016}.

As a result of these aforementioned tradeoffs, there are several notable gaps in the spatio-temporal domains of ecological observations, which lie in the domains defined by high frequency (daily to sub-daily intervals) observations having 1) high to moderate resolutions ($\geq$1 m$^2$ up to 100 ha; Fig. 2A) and 2) decadal or longer durations (Fig. 2B).  Another hole is evident in the high to moderate resolution, large extent (1 million-10 billion ha) domain (Fig. 2C).

>>>>

\author
{Lyndon Estes$^{\ast1, 2}$, Paul R. Elsen$^{3, 4}$, Tim Treuer$^{4}$, Labeeb Ahmed$^{5}$, \\
Kelly Caylor$^{6, 7}$, Jason Chang$^{5}$, Jonathan J. Choi$^{5}$, and Erle Ellis$^{6}$ \\
\\
\normalsize{$^{1}$Graduate School of Geography, Clark University, Worcester MA 01610, USA}\\
\normalsize{$^{2}$Woodrow Wilson School, Princeton University, Princeton, NJ 08544, USA}\\
\normalsize{$^{3}$Department of Environmental Science, Policy, and Management,}\\
\normalsize{University of California Berkeley, Berkeley, CA 94720, USA}\\
\normalsize{$^{4}$Ecology and Evolutionary Biology, Princeton University, Princeton, NJ 08544, USA}\\
\normalsize{$^{5}$Geography and Environmental Systems, University of Maryland Baltimore County,}\\ 
\normalsize{Baltimore, MD 21250, USA}\\
\normalsize{$^{6}$Department of Geography, University of California Santa Barbara, Santa Barbara, CA 93106, USA}\\
\normalsize{$^{7}$Bren School of Environmental Science and Management,}\\
\normalsize{University of California Santa Barbara, Santa Barbara, CA 93106, USA}\\
\normalsize{$^\ast$To whom correspondence should be addressed; E-mail:  LEstes@clarku.edu.}}
% Include the date command, but leave its argument blank.

@@

\linenumbers
\vspace{10pt}
The scales at which ecosystems are observed play a critical role in shaping our understanding of their structure and function \cite{levin_problem_1992,chave_problem_2013,wiens_spatial_1989}.  Ecological patterns emerge from temporal and spatial domains that may be coarser or finer than the processes that shape them, which means that investigation across multiple scales is essential for understanding ecological phenomena \cite{levin_problem_1992,sandel_scale_2009}. This awareness has grown rapidly since the 1980s, accelerated by the need to understand how changes in the global climate, ocean, and land systems are affecting everything from individual populations \cite{tingley_push_2012} to entire biomes \cite{xiao_photosynthetic_2004}, while technological advances in areas such as remote sensing and genetics are making it ever-easier to quantify ecological features across a broad and increasing range of scales \cite{schneider_rise_2001, chave_problem_2013}.

Given the growing awareness of scale, expanding data gathering capabilities, and the fact that the most comprehensive (and arguably best-known) meta-analyses \cite{tilman_ecological_1989,kareiva_spatial_1988} of ecological research scales were published nearly 30 years ago (but see \cite{porter_crop_2005,sandel_scale_2009} for more recent reviews), it is both timely and important to assess the scales of contemporary ecological investigation. To address this need, we quantified the spatial and temporal domains of empirical observations that were reported within recently (2004-2014) published ecological studies. We define domain as the distribution of observations within the spectrum of one or more scale dimensions\footnote{This definition differs slightly from Wiens' \cite{wiens_spatial_1989}, who defined ``domain of scale'' as ``a portion of the scale spectrum within which process-pattern relationships are consistent regardless of scale.''}, and empirical observations as ecological observations collected under un-controlled or non-manipulated conditions. Empirical observations are critical for developing and testing the models that explain why ecological patterns vary in time and space \cite{levin_problem_1992, tilman_ecological_1989}, therefore the spatio-temporal domains of observations provide an important indicator of the field's progress towards achieving a holistic, predictive understanding of ecosystems \cite{chave_problem_2013,levin_problem_1992}.

@@

Contrasting observational dimensions reveals that larger extents are associated with larger spatial replicates (Fig. 2C), while longer durations tend to be associated with longer intervals (Fig. 2B). The latter association primarily reflects a cost-imposed tradeoff between sampling frequency and temporal duration that is characteristic of traditional field-based observation, the dominant observational mode, but is also reflected by the relative domain locations of the four main observational methods, suggesting that high observational frequency poses a durational cost across observational methods. A similar cost tradeoff is also evident in the inverse relationship between resolution and interval that dominates that domain space (Fig. 2A), which again primarily relates to characteristics of field-based observations, in which larger spatial replicates demand greater effort that in turn reduces sampling frequency \cite{kareiva_spatial_1988}. Less obvious is the opposite tradeoff that occurs with remote observation (but shown in Fig. S2 in SI), where finer resolution is desired to increase detail, but typically comes at the cost of longer intervals \cite{estes_platform_2016}.

As a result of these tradeoffs, there are several notable gaps in the spatio-temporal domains of ecological observations, which lie in the domains defined by high frequency (daily to sub-daily intervals) observations having 1) high to moderate resolutions ($\geq$1 m$^2$ up to 100 ha; Fig. 2A) and 2) decadal or longer durations (Fig. 2B).  Another hole is evident in the high to moderate resolution, large extent (1 million-10 billion ha) domain (Fig. 2C).
